Identity of fish fry from the “Tetí” fishery in Eastern Cuba

### 2021-09-06 analysis/paper/paper.Rmd
pdf done
paperLastVersion092021

---
title: 'The Truth is Revealed: Identity of the Fish Fry "Teti" Fishery in Eastern Cuba'
author:
  - Sheila Rodriguez-Machado:
      email: srodr62@lsu.edu
      institute: [UofO]
      correspondence: true
  - Tomas M. Rodriguez-Cabrera:
      email: tomasmichelrodriguez@gmail.com
      institute: [UofA]
      correspondence: false
institute:
  - UofO: Museum of Natural Sciences, Louisiana State University
  - UofA: Sociedad Cubana de Zoologia
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: '../templates/template.docx'
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: '../templates/journal-of-archaeological-science.csl'
abstract: |
  The lack of diagnostic characters to distinguish among species and even genera of gobies (Gobiidae) is still a challenge for science in the 21st century. However, the use of molecular tools applied to species-level identification has brought some light into the taxonomy and systematics of these fish taxa. The fish fry fishery in Eastern Cuba known as “Teti” represents a local fishery and there is no information about the exploitation of this natural resource outside this region. Fishing “Teti” has been a common practice for several years but an exponential increase of tourism in Eastern Cuba represents an additional threat to our biodiversity. Besides, if “Teti” refers to a single species or a species complex is still a knowledge gap in Cuba. The present study aims to identify the Fish Fry "Teti" Fishery in Eastern Cuba using molecular markers. We conducted a barcoding analysis comparing Cuban samples of gobies with samples of the genera *Sicydium* and *Awaous* available in GenBank. Here, we provide evidence that Busck´s Stone-Biting Goby, *Sicydium buscki*, is a component of the fish fry “Teti” fishery in Eastern Cuba. However, whether this fishery includes a single species or more than one species will remain as a mystery until more studies are conducted. Although this study represents the first effort toward and effective management, more thorough investigations and identification of the species comprising the “Teti” fishery are needed. It is imperative to include them in local management plans and to avoid overfishing this resource.  
keywords: |
  amphidromy; Busck's Stone-Biting Goby; COI; DNA barcoding; Gobiidae

@@

```{r setup, echo = FALSE}    
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

#knitr::opts_knit$set(root.dir = '/rrcompendium/')

```{r}

# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("BiocManager", "raster", "ggplot2", "ggthemes", "mapproj", "here",  
             "rentrez", "ape", "seqinr", "phangorn")

package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE, include = FALSE, warn.conflicts = FALSE, quietly = TRUE)
    }
  }
)

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

@@

The family Gobiidae includes almost 2,000 species of marine, freshwater, and brackish fishes worldwide [@eschmeyer2020catalog]. This group is mostly represented by small-sized and bottom-dweller fishes found in tropical and subtropical waters. Such a small size along with cryptic body colors and the lack of diagnostic characters to distinguish among species and even genera is still a challenge for science in the 21st century. Only in the last ten years, more than a hundred gobiid species were described [@eschmeyer2020catalog]. However, the use of molecular tools applied to species-level identification has brought some light into the taxonomy and systematics of these fish taxa [@thacker2011phylogeny; @victor2014three; @keith2017schismatogobius].

Although most species in the family Gobiidae inhabit marine ecosystems, most species inhabiting freshwater or estuarine ecosystems display an amphidromous lifecycle [@myers1949usage; @bell1999overview]. This means that adults spawn in freshwater and when the eggs hatch, they are taken downstream by river flow into the ocean. Then, after spending from a few weeks to a few months (depending on the species), postlarvae return to rivers. This moment of postlarval returning is particularly fascinating because million of them gather at the river mouth as they start going upstream. This migratory behavior, particularly, when the postlarvae are going back to rivers, supports an artisanal fishery around the world [@bell1999overview; @mcdowall2007amphidromy]. Known as goby-fry fisheries, this is a common practice in several countries in the Indo-Pacific [@asis2013first], the Pacific [@castellanos2011goby], and the Caribbean [@neal2009factors]. The problem with it is that during the last years insufficient management of these fisheries has led to a considerable reduction in catches [@bell1999overview]. In addition, information on the identity of the species comprising these goby-fry fisheries in many places is still incomplete [@castellanos2011goby; @asis2013first].

Gobies have been poorly studied in Cuba and available information on their identification and distribution is not conclusive [@lara2010dna]. The fish fry fishery in Eastern Cuba known as “Teti” represents a local fishery and there is no information about the exploitation of this natural resource outside this region. Base on other reports in the Caribbean region [@neal2009factors], some authors claimed that the gobies *Sicydium plumieri* and *Sicydium punctatum* comprise the “Teti” fisheries in Cuba, but there is no solid evidence on that yet. Fishing “teti” has been a common practice for several years but an exponential increase of tourism in Eastern Cuba, due to recent economic policy changes, represents an additional threat to our biodiversity. Besides, if “teti” refers to a single species or a species complex is still a knowledge gap in Cuba. The present study aims to identify the Fish Fry "Teti" Fishery in Eastern Cuba using molecular markers.

@@

Tetis (19–28 mm standard length) were collected 400 m from the river mouth of Río Yumurí (Figure \@ref(fig:mainMap)), eastern Cuba, in December, 2018. Samples were collected using hand nets (1 mm mesh size) 2 m from the riverbank. Specimens were anesthetized using MS 222 (Tricaine Methanesulphonate) and immediately preserved in 80% ethanol. Voucher specimens were deposited in the collection of the Acuario Nacional de Cuba (catalog number: ANC-xxxx).

```{r mainMap, fig.cap="Sampling site at Yumuri river mouth, Eastern Cuba"}

# getting the map of Cuba
mydata <- getData("GADM", country = "cuba", level = 1)  # package raster
mymap <- fortify(mydata)                                # package ggplot2

# sampling site
dots <- data.frame(
  long = c(-74.293460),
  lat = c(20.298537),
  names = c("Yumuri river mouth"),
  stringsAsFactors = FALSE
  )

# main map
mainMap <-  ggplot() +
       geom_blank(data = mymap, aes(x = long, y = lat)) +
       geom_map(data = mymap, map = mymap, aes(group = group, map_id = id), fill = "gray90", color = "black", size = 0.3) +
       scale_x_continuous(limits = c(-74.51, -74.11), expand = c(0, 0)) +
       scale_y_continuous(limits = c(20.13, 20.42), expand = c(0, 0)) +
       geom_point(data = dots, aes(x = long, y = lat), color = "red", size = 4) +     
       coord_equal() +
       theme_bw()

# small square framing the zomm area
zoomOut <- data.frame(long = c(-74.49, -74.49, -74.09, -74.09, -74.49),
                   lat = c(20.13, 20.40, 20.40, 20.13, 20.13))

# inset map
insetMap <- ggplotGrob(
      ggplot() +
      geom_blank(data = mymap, aes(x = long, y = lat)) +
      geom_map(data = mymap, map = mymap, aes(group = group, map_id = id), fill = "gray95", color = "black", size = 0.3) +
      geom_path(data = zoomOut, aes(x = long, y = lat), size = 0.3) +
      scale_x_continuous(limits = c(-85.47, -73.65), expand = c(0, 0)) +
      scale_y_continuous(limits = c(19.5, 23.5), expand = c(0, 0)) +
      coord_map("polyconic") +
      theme_map() +     # package ggthemes
      theme(panel.background = element_rect(fill = NULL)) # fill = NULL or transparent, with color=NA both the background and the frame are transparent
      )

mainMap +
      annotation_custom(grob = insetMap, xmin = -74.12, xmax = -74.37,
                        ymin = 20.31, ymax = 20.42)

@@

Total DNA was extracted from caudal peduncle muscle tissues using the DNAeasy tissue kit (QIAGEN). Partial sequences of the mitochondrial gene cytochrome oxidase I (COI) were amplified through the polymerase chain reaction in 20 μL of reaction volume (one unit of GoTaq DNA polymerase-Promega, 0.2 nM of each primer, 0.2 μM dNTPs, and 1.5 mM MgCl2) [@lara2010dna]. We used the primers COIf (5´-AAYCAYAAAGAYGGYACCCT-3´) and COIr (3´-CCTCNGGRTGNCCRAAGAAYCA-5´) [@palumbi1991simple].

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious], and submitted to GenBank database (accession numbers #####). We used sequences of the genus *Sicydium* as reference for recovering the taxonomic identity of the "Tetis". Sequences of *Awaous banana* and *A. ocellaris* were used as outgroup. All non-Cuban sequences were recovered from GenBank using the R package *rentrez* [@winter2017rentrez]. Duplicated sequences and sequences with missing data were excluded. Sequences were edited using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2015package], and *Biostrings* [@pages2013package]. Final alignment was obtained with the algorithm ClustalOmega [@sievers2011fast] using the R package *msa* [@bodenhofer2015msa].

```{r gettingAPI}

                                                            # DOWNLOADING SEQS FROM GENBANK
# rentrez: this package provides an R interface NCBI API and allows to search databases like GenBank (http://www.ncbi.nih.gov/genbank/)

Sicydium_COI <- "Sicydium[Organism] AND COI[Gene]"

# I set retmax=580 because after running this by default, I got a message with the number of hits (580 in this case)
Sicydium_COI_search <- entrez_search(db = "nuccore", term = Sicydium_COI, retmax = 580)

# given that the search is too big for the system, it needs to be split into two smaller ones
Sicydium_COI_seqs1 <- entrez_fetch(db = "nuccore", id = Sicydium_COI_search$ids[0:295], rettype = "fasta")
Sicydium_COI_seqs2 <- entrez_fetch(db = "nuccore", id = Sicydium_COI_search$ids[296:580], rettype = "fasta")

# GETTING OUTGROUP SEQUENCES OF AWAOUS BANANA (2) AND A. OCELLARIS (2)

Awaous_bananaCOI <- "Awaous_banana[Organism] AND COI[Gene]"
Awaous_bananaCOI_search <- entrez_search(db = "nuccore", term = Awaous_bananaCOI, retmax = 2)
Awaous_bananaCOI_seqs <- entrez_fetch(db = "nuccore", id = Awaous_bananaCOI_search$ids, rettype = "fasta")

Awaous_ocellarisCOI <- "Awaous_ocellaris[Organism] AND COI[Gene]"
Awaous_ocellarisCOI_search <- entrez_search(db = "nuccore", term = Awaous_ocellarisCOI, retmax = 4)
Awaous_ocellarisCOI_seqs <- entrez_fetch(db = "nuccore", id = Awaous_ocellarisCOI_search$ids, rettype = "fasta")

# storing ALL sequences in a single fasta file
write(Sicydium_COI_seqs1, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n")
# 'append = TRUE', I can keep adding seqs
write(Sicydium_COI_seqs2, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)   
write(Awaous_bananaCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)
write(Awaous_ocellarisCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)

@@

```{r editingAPI}

# seqinr: package for sequence manipulation 
# reading the file and renaming the sequences

Sicydium_COI_seqinrFormat <- read.fasta(here(file ="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

Sicydium_COI_names <- attr(Sicydium_COI_seqinrFormat, "name")      # editing species name
Sicydium_COI_names <- gsub("\\..*", "", Sicydium_COI_names)        # eliminating characters after "." using gsub
Sicydium_COI_names <- gsub("^.*\\|", "", Sicydium_COI_names)       # eliminating characters before "|" using gsub

# using library "ape"
Sicydium_COI_apeFormat <- read.GenBank(Sicydium_COI_names)         # this line takes a few minutes to run

# creating a vector (to be added as name of the seqs) with the species, GenBank accession number, and type of gene
Sicydium_COI_GenBank_Ids <- paste(attr(Sicydium_COI_apeFormat, "species"), names(Sicydium_COI_apeFormat), sep = "_COI_")

# writing a new fasta file (2nd) using the vector above 
write.fasta(sequences = Sicydium_COI_seqinrFormat, names = Sicydium_COI_GenBank_Ids, nbchar = 10,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"))

                                     # EDITTING SEQS AND GETTING THEM ALL READY FOR THE PHYLOGENETIC ANALYSIS

### REMOVING sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"),
                        seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

fastaNames <- strsplit(names(nameFasta), '_')                  # creating a vector splitting species names
completeSeqs <- which(! grepl("sp.", fastaNames))              # selecting only those with specific name in them

# writing a new fasta (3rd) file with only those seqs with full scientific names
write.fasta(sequences = nameFasta[completeSeqs], names = names(nameFasta)[completeSeqs],
            here(file.out = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"))

### REMOVING sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"),
                          seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)     # 586 seqs
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]
write.fasta(sequences = fastaLength, names = names(fastaLength),
            here(file.out = "analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"))             # 401 seqs

### Removing DUPLICATE sequences
# using "seqinr" to do it 
# for not using attributes to name sequences
duplicatesFasta <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"), set.attributes = FALSE)

seqnames <- names(duplicatesFasta)
dups <- grep(TRUE,duplicated(duplicatesFasta))
uniqueNames <- seqnames[-dups]
uniqueSeqs <- duplicatesFasta[-dups] 
write.fasta(sequences = uniqueSeqs, names = uniqueNames,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"))               # 204 seqs

# adding samples from Cuba (n=6)
mySamples <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_Cuba.fas"))
write.fasta(mySamples, names(mySamples), here("analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"), open = "a")              # 210 seqs

@@

```{r alignment}
invisible({capture.output({                ### ALIGNING the sequences        # libraries "msa" and "Biostrings"

alignmentFasta <- readDNAStringSet(here("analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"))
set.seed(123)
ClustalOAlignment <- msa(alignmentFasta, "ClustalOmega")
# print(ClustalOAlignment, showConsensus=FALSE, showNames=TRUE, show="complete")

# Checking for the position of sequences that were not edited properly so they still have nucleotides different that A, G, C, T 
nucType <- alphabetFrequency(ClustalOAlignment)
unknownNuc <- nucType[, c(5:14)]  # this subset only contains nucleotides that are not A, G, T, C, or N or gaps
unknownPosition <- which(unknownNuc > 0, arr.ind=T)

# REMOVING samples with too many GAPS at the beginning and the end of the sequences and UNKNOWN nucleotides (ex.: W, K, S) 
colmask(ClustalOAlignment) <- IRanges(start=c(1,666),end=c(25,679))
rowmask(ClustalOAlignment) <- IRanges(start=c(1),end=c(1))
ClustalOAlignment2 <- as(ClustalOAlignment, "DNAStringSet")
writeXStringSet(ClustalOAlignment2, here(file="analysis/data/raw_data/Sicydium_COI_6-alignment.fasta"))            # 209 seqs -- 640 bp

### Replacing gaps at the begining and end with Ns
gapsFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_6-alignment.fasta"), seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE, apply.mask = FALSE)
#getLength(gapsFasta)
replacingGaps <- lapply(gapsFasta, function(x) gsub("-", "N", x))
write.fasta(replacingGaps, names(replacingGaps), here("analysis/data/raw_data/Sicydium_COI_7-gaps.fasta"))

})})

@@

```{r editingTree, echo = FALSE}

# Editing and visualizing the tree to represent --- library: ggtree
tree <- read.tree(here(file="analysis/data/derived_data/Sicydium_ML_tree.tre"))

# checking that the tree is rooted
rootedTree <- is.rooted(tree)   # package ape. showing the tree root: geom_rootpoint(color="pink", size=5) 
# getting the node number for each main clade. This info is important because those #s are the one needed to edit each main clade/species
set.seed(123)
nodeNumber <- ggtree(tree) + geom_text(aes(label=node), hjust=-.5, col = "blue", size = 1) # looks horrible but it gives the clade number

>>>>

---
title: 'Identity of fish fry from the “Tetí” fishery in Eastern Cuba'
author:
  - Sheila Rodriguez-Machado:
      email: srodr62@lsu.edu
      institute: [UofO]
      correspondence: true
  - Tomas M. Rodriguez-Cabrera:
      email: tomasmichelrodriguez@gmail.com
      institute: [UofA]
      correspondence: false
  - Prosanta Chakrabarty:
      email: prosanta@lsu.edu
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Museum of Natural Sciences, Louisiana State University
  - UofA: Sociedad Cubana de Zoologia
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: '../templates/template.docx'
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: '../templates/cell-structure-and-function.csl'
abstract: |
  Fisheries of juvenile and larval fishes (i.e., fish-fry fisheries) often exploit the migratory behavior of diadromous species on their return to rivers. Although these fisheries are common worldwide, fish-fry fisheries are subject to insufficient management and the identity of the target species is frequently unknown. The artisanal fish-fry fishery in Cuba is known as “Tetí”, and is restricted to the eastern region of the country. Fishing “Tetí” has been practiced for hundreds of years, but an increase in tourism has elevated demand in recent years and may pose a threat to the conservation of this natural resource. It remains unclear if “Tetí” refers to one or multiple species. The present study aims to identify the fish fry that make up the “Tetí” fishery in Eastern Cuba with genetic data. We first used a standard nucleotide BLAST to identify the sequences. Then, we recovered the phylogenetic position of these samples based on a maximum likelihood phylogenetic analysis that also used available sequences from GenBank. Here, we provide evidence that Busck´s Stone-Biting Goby, Sicydium buscki, is a component of the “Tetí” fishery in Cuba. However, whether this fishery includes one or more species remains a mystery until additional samples can be analyzed. However, with this species identification the first efforts towards effective management can be taken to manage this fishery. Collecting additional scientific and fishing data from the “Tetí” fishery and including them in local management plans will be an important strategy for avoiding overfishing this resource.  
keywords: |
  amphidromy; Busck's Stone-Biting Goby; COI; DNA barcoding; Gobiidae

@@

```{r setup, echo = FALSE}    
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
#knitr::opts_knit$set(root.dir = '/rrcompendium/')
```

```{r}
# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("BiocManager", "raster", "ggplot2", "ggthemes", "mapproj", "here",  
             "rentrez", "ape", "seqinr", "phangorn",
             "adegenet", "pegas")
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE, include = FALSE, warn.conflicts = FALSE, quietly = TRUE)
    }
  }
)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

@@

The family Gobiidae includes almost 2,000 species of marine, freshwater, and brackish water fishes worldwide [@eschmeyer2020catalog]. This group is mostly represented by small and bottom-dwelling fishes found in tropical and subtropical waters. Their small size along with cryptic body colors and the paucity of clear diagnostic characters to distinguish among taxa is a substantial challenge. However, the use of molecular tools applied to species-level identification has shed light on the taxonomy of these fishes [@keith2011phylogeny; @thacker2011phylogeny; @victor2014three; @kuang2018phylogenomic] and in the last ten years, more than one hundred gobiid species have been described [@eschmeyer2020catalog].

Although most gobies inhabit marine ecosystems, some species inhabiting freshwater or estuarine ecosystems display an amphidromous lifecycle [@myers1949usage; @bell1999overview]: meaning that adults spawn in freshwater and the post-hatch larvae are carried downstream by rivers into the ocean. After spending weeks to months (depending on the species) in these marine habitats, postlarval stage individuals return to rivers. This period of postlarval return is particularly fascinating because million of them gather at the river mouth before starting their journey upstream. This migratory behavior supports artisanal fisheries worldwide [@bell1999overview; @mcdowall2007amphidromy]. Goby-fry fisheries are common practice in several countries in the Indo-Pacific [@asis2013first], the Pacific [@castellanos2011goby], and the Caribbean (@erdman1961notes; @kwaketal2016). One problem with this style of fishery is that insufficient management has led to a considerable reduction in catch [@bell1999overview]. In addition, information on the identity of the species comprising these goby-fry fisheries is still incomplete in many places [@castellanos2011goby; @asis2013first].

Gobies have been poorly studied in Cuba and available information on their identification and distribution is not conclusive [@lara2010dna]. The fish-fry “Tetí” fishery in Eastern Cuba represents a local artisanal fishery and there is no scientific information about the exploitation of this natural resource outside of the region. Based on other reports from the Caribbean (*e.g.*, @watson2000), some authors have claimed that the gobies *Sicydium plumieri* and *Sicydium punctatum* make up the “Tetí” fishery in Cuba, but evidence for those identications are lacking. “Tetí” has been historically exploited as the main ingredient of “traditional and exotic” dishes [@pichardoytapia1862; A. Hartman, pers. comm., August 01, 2011). However, recent changes to Cuban economic policies have promoted considerable growth in tourism [@rutty2019tourism], leading to increased fisheries pressure. Annual data series show increases in the number of visitors and infrastructure development in Baracoa [@baracoaStats2015; @baracoaStats2019], where this fishery takes place. Such actions along with the lack of fishing statistics might represent a threat to the conservation of these fish. This study aims to identify fish-fry species involved in the “Tetí” fishery in Eastern Cuba using molecular data.

@@

Nine “Tetís” (19–28 mm standard length) (Figure \@ref(fig:tetiPhoto)) were collected 400 m from the river mouth of the Río Yumurí (Figure \@ref(fig:mainMap)), eastern Cuba, in December 2018. Specimens were collected using hand nets (1 mm mesh size) 2 m from the riverbank, anesthetized using MS 222 (Tricaine methanesulphonate), and immediately preserved in 80% ethanol. Voucher specimens were deposited in the collection of the Acuario Nacional de Cuba (catalog number: ANC-12.2.454).

```{r tetiPhoto, fig.cap="Lateral view of goby fry “Tetí” captured in Río Yumurí, Cuba. Photo by Tomás M. Rodríguez-Cabrera."}

@@

```{r mainMap, fig.cap="Map of Cuba indicating where the fish-fry “Tetí” fishery has been reported. The outer circle represents the sampling locality of this study."}

# getting the map of Cuba
mydata <- getData("GADM", country = "cuba", level = 1)  # package raster
mymap <- fortify(mydata)                                # package ggplot2
# sampling site
dots <- data.frame(
  long = c(-74.293460),
  lat = c(20.298537),
  names = c("Yumuri river mouth"),
  stringsAsFactors = FALSE
  )

# main map
mainMap <-  ggplot() +
       geom_blank(data = mymap, aes(x = long, y = lat)) +
       geom_map(data = mymap, map = mymap, aes(group = group, map_id = id), fill = "gray90", color = "black", size = 0.3) +
       scale_x_continuous(limits = c(-74.51, -74.11), expand = c(0, 0)) +
       scale_y_continuous(limits = c(20.13, 20.42), expand = c(0, 0)) +
       geom_point(data = dots, aes(x = long, y = lat), color = "red", size = 4) +     
       coord_equal() +
       theme_bw()
# small square framing the zomm area
zoomOut <- data.frame(long = c(-74.49, -74.49, -74.09, -74.09, -74.49),
                   lat = c(20.13, 20.40, 20.40, 20.13, 20.13))
# inset map
insetMap <- ggplotGrob(
      ggplot() +
      geom_blank(data = mymap, aes(x = long, y = lat)) +
      geom_map(data = mymap, map = mymap, aes(group = group, map_id = id), fill = "gray95", color = "black", size = 0.3) +
      geom_path(data = zoomOut, aes(x = long, y = lat), size = 0.3) +
      scale_x_continuous(limits = c(-85.47, -73.65), expand = c(0, 0)) +
      scale_y_continuous(limits = c(19.5, 23.5), expand = c(0, 0)) +
      coord_map("polyconic") +
      theme_map() +     # package ggthemes
      theme(panel.background = element_rect(fill = NULL)) # fill = NULL or transparent, with color=NA both the background and the frame are transparent
      )
mainMap +
      annotation_custom(grob = insetMap, xmin = -74.12, xmax = -74.37,
                        ymin = 20.31, ymax = 20.42)

@@

Total DNA was extracted from caudal peduncle muscle tissues using the DNeasy tissue kit (QIAGEN). Partial sequences of the mitochondrial gene cytochrome oxidase I (COI) were amplified through the polymerase chain reaction in 20 μL of reaction volume (one unit of GoTaq DNA polymerase-Promega, 0.2 nM of each primer, 0.2 μM dNTPs, and 1.5 mM MgCl2) [@lara2010dna]. We used the primers COIf (5´-AAYCAYAAAGAYGGYACCCT-3´) and COIr (3´-CCTCNGGRTGNCCRAAGAAYCA-5´) [@palumbi1991simple].

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious] and submitted to GenBank (accession numbers: MZ848370– MZ848378). We used a standard nucleotide BLAST (BLASTN 2.11.0; @zhangetal2000), optimized for highly similar sequences (MegaBLAST; @morgulis2008database) (Query IDs: 1237–1245), and recovered the identity of the Cuban sequences with 99–100% identity.

We also aimed to recover the phylogenetic position of the fish-fry samples. For this analysis, all non-Cuban sequences were retrieved from GenBank using the R package *rentrez* [@winter2017rentrez] (Appendix 1). Duplicated sequences and sequences with missing data were excluded. The dataset was edited and aligned using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2015package], and *Biostrings* [@pages2013package]. The final alignment was obtained with the algorithm ClustalOmega [@sievers2011fast] using the R package *msa* [@bodenhofer2015msa].

```{r gettingAPI}
                                                            # DOWNLOADING SEQS FROM GENBANK
# rentrez: this package provides an R interface NCBI API and allows to search databases like GenBank (http://www.ncbi.nih.gov/genbank/) 
Sicydium_COI <- "Sicydium[Organism] AND COI[Gene]"
# I set retmax=580 because after running this by default, I got a message with the number of hits (580 in this case)
Sicydium_COI_search <- entrez_search(db = "nuccore", term = Sicydium_COI, retmax = 580)
# given that the search is too big for the system, it needs to be split into two smaller ones
Sicydium_COI_seqs1 <- entrez_fetch(db = "nuccore", id = Sicydium_COI_search$ids[0:295], rettype = "fasta")
Sicydium_COI_seqs2 <- entrez_fetch(db = "nuccore", id = Sicydium_COI_search$ids[296:580], rettype = "fasta")
# GETTING OUTGROUP SEQUENCES OF AWAOUS BANANA (2) AND A. OCELLARIS (2)
Awaous_bananaCOI <- "Awaous_banana[Organism] AND COI[Gene]"
Awaous_bananaCOI_search <- entrez_search(db = "nuccore", term = Awaous_bananaCOI, retmax = 2)
Awaous_bananaCOI_seqs <- entrez_fetch(db = "nuccore", id = Awaous_bananaCOI_search$ids, rettype = "fasta")
Awaous_ocellarisCOI <- "Awaous_ocellaris[Organism] AND COI[Gene]"
Awaous_ocellarisCOI_search <- entrez_search(db = "nuccore", term = Awaous_ocellarisCOI, retmax = 4)
Awaous_ocellarisCOI_seqs <- entrez_fetch(db = "nuccore", id = Awaous_ocellarisCOI_search$ids, rettype = "fasta")
# storing ALL sequences in a single fasta file
write(Sicydium_COI_seqs1, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n")
# 'append = TRUE', I can keep adding seqs
write(Sicydium_COI_seqs2, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)   
write(Awaous_bananaCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)
write(Awaous_ocellarisCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)

@@

```{r editingAPI}
# seqinr: package for sequence manipulation 
# reading the file and renaming the sequences
Sicydium_COI_seqinrFormat <- read.fasta(here(file ="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)
Sicydium_COI_names <- attr(Sicydium_COI_seqinrFormat, "name")      # editing species name
Sicydium_COI_names <- gsub("\\..*", "", Sicydium_COI_names)        # eliminating characters after "." using gsub
Sicydium_COI_names <- gsub("^.*\\|", "", Sicydium_COI_names)       # eliminating characters before "|" using gsub
# using library "ape"
Sicydium_COI_apeFormat <- read.GenBank(Sicydium_COI_names)         # this line takes a few minutes to run
# creating a vector (to be added as name of the seqs) with the species, GenBank accession number, and type of gene
Sicydium_COI_GenBank_Ids <- paste(attr(Sicydium_COI_apeFormat, "species"), names(Sicydium_COI_apeFormat), sep = "_COI_")
# writing a new fasta file (2nd) using the vector above 
write.fasta(sequences = Sicydium_COI_seqinrFormat, names = Sicydium_COI_GenBank_Ids, nbchar = 10,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"))
                                     # EDITTING SEQS AND GETTING THEM ALL READY FOR THE PHYLOGENETIC ANALYSIS
### REMOVING sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"),
                        seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)
fastaNames <- strsplit(names(nameFasta), '_')                  # creating a vector splitting species names
completeSeqs <- which(! grepl("sp.", fastaNames))              # selecting only those with specific name in them  
# writing a new fasta (3rd) file with only those seqs with full scientific names
write.fasta(sequences = nameFasta[completeSeqs], names = names(nameFasta)[completeSeqs],
            here(file.out = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"))
### REMOVING sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"),
                          seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)     # 586 seqs
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]
write.fasta(sequences = fastaLength, names = names(fastaLength),
            here(file.out = "analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"))             # 401 seqs
### Removing DUPLICATE sequences
# using "seqinr" to do it 
# for not using attributes to name sequences
duplicatesFasta <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"), set.attributes = FALSE) 
seqnames <- names(duplicatesFasta)
dups <- grep(TRUE,duplicated(duplicatesFasta))
uniqueNames <- seqnames[-dups]
uniqueSeqs <- duplicatesFasta[-dups] 
write.fasta(sequences = uniqueSeqs, names = uniqueNames,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"))               # 204 seqs
# adding samples from Cuba (n=9)
mySamples <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_Cuba_n=9.fas"))
write.fasta(mySamples, names(mySamples), here("analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"), open = "a")              # 210 seqs

@@

```{r editingTree, echo = FALSE}
# Editing and visualizing the tree to represent --- library: ggtree
tree <- read.tree(here(file="analysis/data/derived_data/Sicydium_ML_tree.tre"))
# checking that the tree is rooted
rootedTree <- is.rooted(tree)   # package ape. showing the tree root: geom_rootpoint(color="pink", size=5) 
# getting the node number for each main clade. This info is important because those #s are the one needed to edit each main clade/species
set.seed(123)
nodeNumber <- ggtree(tree) + geom_text(aes(label=node), hjust=-.5, col = "blue", size = 1) # looks horrible but it gives the node number

```{r genetic diversity, echo = FALSE}
# uses packages "pegas", which depends on "adegenet"

diversityData <- read.dna(here("analysis/data/raw_data/Sicydium_COI_Cuba_n=9.fas"), format = "fasta")
diversityToUse <- haplotype(diversityData)
